---
title: "R Project: National Parks Species"
---

# libraries

```{r}
libs <- c("tidyverse", "tidytuesdayR", "scales", "dplyr", "stringr", "ggthemes", "RColorBrewer", "viridis")

installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == F)) {
    install.packages(libs[!installed_libs])
} else{
  print("All the libraries already installed")
}
```

```{r}
library(tidyverse)
library(tidytuesdayR)
library(scales)
library(dplyr)
library(stringr)
library(ggthemes)
library(RColorBrewer)
```

```{r message=FALSE, warning=FALSE}
data <- read_csv("species.csv")

data

```

## Amount of different species in the USA National Parks

```{r}
# Amount of species in every park
species_amount <- 
  data %>% 
    filter(Occurrence %in% c("Present", "Probably Present")) %>%   
    group_by(ParkName) %>% 
    distinct(SciName, .keep_all = TRUE) %>%
    ungroup()

species_amount

barchart_species_amount <-
  ggplot(data = species_amount, aes(x = ParkCode, fill = CategoryName, colour = CategoryName)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Amount of different species in the USA National Parks",
    x = "National Park Code",
    y = "Number of Species",
    fill = "Category Name"
  ) +
  guides(
    color = "none"
  ) +
  scale_fill_viridis_d(option = "viridis")

barchart_species_amount
```

## Observations by Species grouped in Family

```{r message=FALSE, warning=FALSE}

ggplot(data, aes(x = CategoryName, y = Observations)) +
  geom_jitter(width = 0.3, height = 0, color = "forestgreen", alpha = 0.6) +
  theme_minimal() +
  coord_flip() +
  labs(
    title = "Observations by Species grouped in Family",
    x = "Family",
    y = "Number of Observations"
  )

```

## Proportion of Native vs Non-Native Species

## Native vs Non-Native Species by Park

```{r message=FALSE, warning=FALSE}

nativeness_count <- data %>%
  filter(!is.na(Nativeness)) %>% 
  group_by(Nativeness) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Pie chart
ggplot(nativeness_count, aes(x = "", y = Count, fill = Nativeness)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  ggtitle("Proportion of Native vs Non-Native Species") +
  theme(legend.title = element_blank()) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), color = "white")

ggplot(data %>% filter(!is.na(Nativeness)), aes(x = "", fill = Nativeness)) +
  geom_bar(width = 1, position = "fill") +
  coord_polar(theta = "y") +
  theme_void() +
  facet_wrap(~ ParkCode, scales = "free", ncol = 4) +
  ggtitle("Native vs Non-Native Species by Park")

```

## Average Amount of Different Evidence by Abundance Category

```{r}
abundance_average_evidence <- 
  data %>% 
    filter(!is.na(Abundance)) %>% 
    group_by(Abundance) %>% 
    summarise(
      AverageReferences = mean(References), 
      AverageObservations = mean(Observations),
      AverageVouchers = mean(Vouchers),
      .groups = "drop"
    )

abundance_average_evidence

abundance_average_evidence_long <- abundance_average_evidence %>%
  pivot_longer(cols = starts_with("Average"), 
               names_to = "Variable", 
               values_to = "Value")

barchart_abundance_evidence <- 
  ggplot(abundance_average_evidence_long, aes(x = Abundance, y = Value, fill = Variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(Value, 1)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5,
            hjust = 0.5) +
  scale_fill_discrete(labels = c("AverageReferences" = "References", 
                                "AverageObservations" = "Observations", 
                                "AverageVouchers" = "Vouchers")) + 
  labs(
    title = "Average Amount of Different Evidence by Abundance Category",
    x = "Abundance Category",
    y = "Average Value",
    fill = "Kind of Evidence"
  ) +
  theme_minimal()

barchart_abundance_evidence
```

## Number of Rare Species by National Park (G2, G1, GH)

```{r message=FALSE, warning=FALSE}

rare_animals <- data %>%
  filter(GRank %in% c("G1","G2", "GH")) %>%
  group_by(ParkName) %>%
  summarise(RareSpeciesCount = n()) %>%
  arrange(desc(RareSpeciesCount))

ggplot(rare_animals, aes(x = reorder(ParkName, -RareSpeciesCount), y = RareSpeciesCount)) +
  geom_bar(stat = "identity", fill = "darkred") +
  theme_minimal() +
  ggtitle("Number of Rare Species by National Park (G2, G1, GH)") +
  xlab("National Park") +
  ylab("Count of Rare Species") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

## Proportion of Spiders Feeding Strategies in the Great Smoky Mountains National Park

```{r}
feeding_strategies_proportions <- data %>%
  filter(str_detect(ParkTags, "Habit:") & str_detect(ParkTags, "Arachnid")) %>%
  mutate(Habit = str_extract(ParkTags, "Habit: [^;]+")) %>% # Extract "Habit:" value
  mutate(Habit = str_remove(Habit, "Habit: ")) %>% # Remove "Habit:" prefix
  separate_rows(Habit, sep = "[,\\s]+") %>% # Split on spaces or commas
  group_by(SciName) %>%
  summarise(Habit = paste(Habit, collapse = ", ")) %>% # Combine all habits per SciName
  mutate(FeedingStrategy = case_when(
    str_detect(Habit, "Hunter") & str_detect(Habit, "Web-Builder") ~ "Both",
    str_detect(Habit, "Hunter") ~ "Hunter",
    str_detect(Habit, "Web-Builder") ~ "Web-Builder"
  )) %>%
  filter(!is.na(FeedingStrategy)) %>% 
  group_by(FeedingStrategy) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Proportion = Count / sum(Count))

feeding_strategies_proportions

feeding_strategies_proportions_plot <- ggplot(feeding_strategies_proportions, aes(x = "", y = Proportion, fill = FeedingStrategy)) +
  geom_bar(stat = "identity", alpha = 0.8, width = 1) +
  coord_polar(theta = "y") +
  labs(
    title = "Proportion of Spiders Feeding Strategies \nin the Great Smoky Mountains National Park",
    fill = NULL
  ) +
  scale_fill_brewer(palette = "Set1") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

feeding_strategies_proportions_plot
```

## Amount of Different Species in Locations of Grand Canyon National Park

```{r}
species_by_location <- data %>%
  filter(str_detect(ParkTags, "Geographic Regions:") & ParkCode == "GRCA") %>%
  mutate(Region = str_extract(ParkTags, "Geographic Regions: [^;]+")) %>%
  mutate(Region = str_remove(Region, "Geographic Regions: ")) %>% 
  separate_rows(Region, sep = ",") %>% 
  mutate(Region = str_squish(Region)) %>% 
  group_by(Region) %>%
  summarise(Count = n(), .groups = "drop") %>% 
  arrange(Count) %>%
  mutate(Region = factor(Region, levels = Region))

species_by_location

species_by_location_graph <- species_by_location %>% 
  ggplot(aes(x = Region, y = Count, fill = Region)) +
  geom_col() +
  geom_text(aes(label = Count), vjust = -0.5, size = 3.5) +
  labs(
    title = "Amount of Different Species in Locations of Grand Canyon National Park",
    y = NULL,
    x = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(size = 8.5),
    legend.position = "none"
  ) +
  scale_fill_brewer(palette = "YlGn") 

species_by_location_graph
  
```

## 10 most referenced historic extirpated species in the Great Smoky Mountains National Park

```{r}
most_referenced_historic_extirpated_species <- data %>%
  filter(str_detect(ParkTags, "Park historic") & str_detect(ParkTags, "Park extirpated")) %>% 
  arrange(desc(References)) %>% 
  slice_head(n = 10) %>% 
  mutate(
    CommonNames = str_to_title(CommonNames),
    CommonNames = factor(CommonNames, levels = CommonNames[order(References, decreasing = TRUE)])
    )
most_referenced_historic_extirpated_species

most_referenced_historic_extirpated_species_boxplot <- most_referenced_historic_extirpated_species %>% 
  ggplot(aes(x = References, fill = CommonNames)) +
  geom_bar() +
  labs(
    title = "10 most referenced historic extirpated species\nin the Great Smoky Mountains National Park",
    fill = "Species common names",
    y = NULL
  ) + 
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5)
  )

most_referenced_historic_extirpated_species_boxplot

```

## Clustering of Parks by GRank and SRank

```{r message=FALSE, warning=FALSE}

rank_data <- data %>%
  filter(!is.na(GRank), !is.na(SRank)) %>%
  group_by(ParkCode) %>%
  summarise(GRank_Count = n_distinct(GRank),
            SRank_Count = n_distinct(SRank), .groups = 'drop')


set.seed(123)
rank_data$Cluster <- kmeans(scale(rank_data[, -1]), centers = 4)$cluster

# Cluster visualization
ggplot(rank_data, aes(x = GRank_Count, y = SRank_Count, color = as.factor(Cluster), label = ParkCode)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text(nudge_y = -15, size = 2) +
  labs(title = "Clustering of Parks by GRank and SRank",
       x = "Number of Unique GRanks",
       y = "Number of Unique SRanks",
       color = "Cluster") +
  theme_minimal()

```

## Count of Observations by Top 10 Orders (Most frequent animal orders)

```{r message=FALSE, warning=FALSE}

top_orders <- data %>%
  filter(CategoryName %in% c("Mammal", "Bird", "Reptile", "Amphibian", "Fish"), 
         !is.na(Order)) %>%
  count(Order, sort = TRUE) %>%
  top_n(10, n) %>%
  pull(Order)

filtered_data <- data %>%
  filter(Order %in% top_orders, CategoryName %in% c("Mammal", "Bird", "Reptile", "Amphibian", "Fish"))

ggplot(filtered_data, aes(x = Order, fill = Order)) +
  geom_bar() +
  labs(title = "Count of Observations by Top 10 Orders (Most frequent animal orders)",
       x = "Order",
       y = "Number of Observations",
       fill = "Order") +
  theme_minimal() +
  coord_flip() 

```

## Species Occurence in the parks

```{r message=FALSE, warning=FALSE}

status_counts <- data %>%
  group_by(ParkCode, Occurrence) %>%
  summarize(Count = n(), .groups = 'drop')

ggplot(status_counts, aes(x = ParkCode, y = Count, fill = Occurrence)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_minimal() +
  labs(title = "Species Occurence in the parks",
       x = "Park",
       y = "Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Species Occurence in the parks

```{r message=FALSE, warning=FALSE}

status_counts <- data %>%
  group_by(ParkCode, Abundance) %>%
  summarize(Count = n(), .groups = 'drop')

ggplot(status_counts, aes(x = ParkCode, y = Count, fill = Abundance)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_minimal() +
  labs(title = "Species Occurence in the parks",
       x = "Park",
       y = "Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

## Top Orders by GRank Count per Category

```{r message=FALSE, warning=FALSE}
filtered_data <- data %>%
  filter(GRank %in% c("G2", "G1", "GH"))

# Count GRank occurrences by CategoryName and Order
rank_counts <- filtered_data %>%
  group_by(CategoryName, Order) %>%
  summarise(Count = n(), .groups = "drop")

# Identify the Order with the maximum count for each CategoryName
top_orders <- rank_counts %>%
  group_by(CategoryName) %>%
  slice_max(Count, n = 1, with_ties = FALSE)

# Plot the results
ggplot(top_orders, aes(x = reorder(CategoryName, -Count), y = Count, fill = Order)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top Orders by GRank Count per Category",
    x = "Category Name",
    y = "Count of GRank (G2, G1, GH)",
    fill = "Order"
  ) +
  theme_minimal()
```

## Top 10 Species by References-to-Observations Ratio

```{r message=FALSE, warning=FALSE}

# Calculate References-to-Observations and Vouchers-to-Observations ratios
ratio_data <- filtered_data %>%
  mutate(
    Ref_Obs_Ratio = References / Observations
  ) %>%
  filter(Observations > 0)  # Ensure there are observations to calculate the ratios

# Top 10 species by References-to-Observations ratio
top_ref_obs_ratio <- ratio_data %>%
  arrange(desc(Ref_Obs_Ratio)) %>%
  slice_head(n = 10)

# Bar chart for top 10 species by References-to-Observations ratio
ggplot(top_ref_obs_ratio, aes(x = reorder(CommonNames, Ref_Obs_Ratio), y = Ref_Obs_Ratio)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Species by References-to-Observations Ratio",
    x = "Species (Common Name)",
    y = "References-to-Observations Ratio"
  ) +
  theme_minimal()

```

## Top Orders by GRank Percentage per Category

```{r message=FALSE, warning=FALSE}

filtered_data <- data %>%
  filter(GRank %in% c("G2", "G1", "GH"))


rank_counts <- filtered_data %>%
  group_by(CategoryName, Order) %>%
  summarise(Count = n(), .groups = "drop")


category_totals <- rank_counts %>%
  group_by(CategoryName) %>%
  summarise(Total = sum(Count), .groups = "drop")

category_totals

rank_percentages <- rank_counts %>%
  left_join(category_totals, by = "CategoryName") %>%
  mutate(Percentage = (Count / Total) * 100)


top_orders <- rank_percentages %>%
  group_by(CategoryName) %>%
  slice_max(Percentage, n = 1, with_ties = FALSE)

# Plot the results
ggplot(top_orders, aes(x = reorder(CategoryName, -Percentage), y = Percentage, fill = Order)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top Orders by GRank Percentage per Category",
    x = "Category Name",
    y = "Percentage of GRank (G2, G1, GH)",
    fill = "Order"
  ) +
  theme_minimal()

```

## Amount and Abundance of Cultivated Species in Different Parks

```{r}
# data %>% filter(str_detect(NativenessTags, "Cultivated"))

cultivated_species <- data %>% 
  filter(str_detect(NativenessTags, "Cultivated") & !is.na(Abundance))
invasive_species

cultivated_abundance_by_park_plot <- cultivated_species %>%
  ggplot(aes(x = ParkName, fill = Abundance)) +
  geom_dotplot() +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, 30),
    breaks = seq(0, 30, 5)
  ) +
  labs(
    title = "Amount and Abundance of Cultivated Species in Different Parks",
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

cultivated_abundance_by_park_plot

```

## Relation of Winter, Summer, and Whole Year Birds by Park

```{r}
bird_data <- data %>% 
  filter(CategoryName == "Bird")

season_birds_by_park_data <- bird_data %>%
  mutate(
    NPSTags = gsub(";|,", ",", NPSTags),        # Standardize delimiters to ","
    NPSTags = ifelse(is.na(NPSTags), "Whole Year / Unknown", NPSTags)    # Mark missing tags
  ) %>%
  separate_rows(NPSTags, sep = ",\\s*") %>%     # Split tags into individual rows
  mutate(NPSTags = ifelse(!NPSTags %in% c("Winter", "Summer"), "Whole Year / Unknown", NPSTags)) %>%
  group_by(ParkCode, NPSTags) %>%
  summarise(Count = n(), .groups = "drop")

plot_data <- season_birds_by_park_data %>%
  group_by(ParkCode) %>%
  mutate(Proportion = Count / sum(Count)) %>% 
  filter(!(NPSTags == "Whole Year / Unknown" & Proportion == 1))

season_birds_by_park_plot <-
  ggplot(plot_data, aes(x = "", y = Proportion, fill = NPSTags)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar(theta = "y") +
    facet_wrap(~ParkCode) +
    labs(
      title = "Relation of Winter, Summer, and Whole Year Birds by Park",
      x = NULL,
      y = NULL,
      fill = "Tag"
    ) +
    theme_void() +  # Clean theme for pie charts
    theme(
      strip.text = element_text(size = 12),  # Adjust facet labels
      legend.title = element_blank(),
      legend.position = c(0.95,0.2)
    )

season_birds_by_park_plot
```

## Relation Between Vagrants and Residents by Category

```{r}
vagrant_vs_resident_by_category_data <- data %>%
  mutate(
    NPSTags = gsub(";|,", ",", NPSTags)) %>%
  separate_rows(NPSTags, sep = ",\\s*") %>%     # Split tags into individual rows
  filter(NPSTags %in% c("Vagrant", "Resident")) %>%
  group_by(CategoryName, NPSTags) %>%
  summarise(Count = n(), .groups = "drop")

plot_data <- vagrant_vs_resident_by_category_data %>%
  group_by(CategoryName) %>%
  mutate(Proportion = Count / sum(Count))

vagrant_vs_resident_by_category_plot <-
  ggplot(plot_data, aes(x = "", y = Proportion, fill = NPSTags)) +
    geom_bar(stat = "identity") +
    coord_polar(theta = "y") +
    facet_wrap(~CategoryName) +
    labs(
      title = "Relation Between Vagrants and Residents by Category",
      x = NULL,
      y = NULL,
      fill = NULL
    ) +
    theme_void() +  # Clean theme for pie charts
    theme(
      strip.text = element_text(size = 8.5),  # Adjust facet labels
      plot.title = element_text(hjust = 0.5, margin = margin(b = 10, t = 10)),
      legend.position = c(0.75, 0.2)
    )

vagrant_vs_resident_by_category_plot
```

## Relationship Between Observations and Vouchers

```{r message=FALSE, warning=FALSE}

filtered_data <- data %>%
  filter(Observations > 0) %>%
  filter(Vouchers > 0) %>%
  mutate(
    Relation = case_when(
      Vouchers == Observations ~ "Equal",
      Observations > Vouchers ~ "More Observations",
      TRUE ~ "More Vouchers"
    )
  )

# Create scatterplot with grouped colors and logarithmic axes
ggplot(filtered_data, aes(x = Vouchers, y = Observations, color = Relation)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Relationship Between Observations and Vouchers",
    x = "Number of Vouchers (Log Scale)",
    y = "Number of Observations (Log Scale)",
    color = "Relation"
  ) +
  theme_minimal() +
  scale_color_manual(
    values = c(
      "Equal" = "purple",
      "More Observations" = "green",
      "More Vouchers" = "orange"
    )
  ) +
  scale_x_log10() +  
  scale_y_log10() 
```

## Percentage of Categories: Observations vs Vouchers

```{r message=FALSE, warning=FALSE}

pie_data <- filtered_data %>%
  group_by(Relation) %>%
  summarize(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100,
         Label = paste0(round(Percentage, 1), "%"))  # Create labels with percentages

# Create pie chart with percentage labels
ggplot(pie_data, aes(x = "", y = Percentage, fill = Relation)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5), size = 4) +  # Add percentage labels
  labs(
    title = "Percentage of Categories: Observations vs Vouchers",
    fill = "Relation"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_manual(
    values = c(
      "Equal" = "purple",
      "More Observations" = "green",
      "More Vouchers" = "orange"
    )
  )

```


